Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-20 articles/cycle/response-referee1.tex

    \referee{%
        The point is well taken that trimlines do not necessarily indicate past
        ice surface elevations, but the upper limit of temperate ice with cold
        ice above (pg. 18, line 5 to pg. 19, line 4). It is an important point,
        because model LGM ice surfaces are far above most trimline elevations
        (as noted and in Fig.~6a,b). The pertinent results are shown in
        Fig.~6c, and I agree there is good support for the cold vs. temperate
        (basal) ice hypothesis.}

    Thank you for emphasizing this point.

    \referee{%
        It might help general readers to spell out the interpretation even more
        in the text. That is, as I understand it, the observed trimlines should
        coincide with the boundaries between model areas of frozen vs.
        temperate beds, so the dots in Fig.~6c should all lie on the borders
        between the hatched and white areas. The sentence on p.~19, l.~20-21,
        is confusing in this regard. Incidentally, it would also help to add
        the word "basal" to the last sentence of the Fig.~6c caption:
        "...experienced temperate basal ice for ...".}

    We realise this is confusing. We had chosen this approach because temperate
    basal ice tends to occur above the trimlines during periods of warmer
    climate. To avoid confusion, Fig.~6c has been simplified.

@@

    \referee{%
        One reason for the remaining discrepancies in Fig.~6c could be temporal
        variations in the model boundaries, that are aggregated in time by the
        "< 1\,kyr" criterion for the hatching and the grouping of all trimline
        data. To go into this in more detail, in principle Fig.~6c could be
        expanded to show the model basal frozen-temperate boundaries at
        particular times (21.5, 22.5, etc, ka), with only the dots for each
        time period superimposed. But that may not be worth it unless there are
        large temporal variations in the model boundaries.}

    Fig.~6c has been simplified to show the basal thermal boundary at the age
    of maximum ice thickness. Although this results in a time-transgressive
    picture, it means that the frozen areas indicated on the new figure are
    contemporaneous with the plotted ages and surface topography contours, and
    data shown on panels~a and b.

    Although the model output allows to study the migration of the basal
    thermal boundary over time into more detail, the basal velocity, which
    controls erosion, should probably be considered as well. We leave this for
    future studies.

@@

    \referee{%
        A slight concern is that the majority of the trimline data seems to be
        orange dots i.e., older that 27\,ka in the timescale of Fig.~6c. The
        period for the model hatching extends back only to 29\,ka. Hopefully,
        most of the orange-dotted data are within that period and are not older
        than 29\,ka(?).}

    Indeed, much of the mountainous areas reach maximum ice thickness during
    early MIS~2 when ice is colder and thus stiffer. This is also visible on
    Fig.~5. Some areas even reach maximum ice thickness during MIS~4 when the
    bedrock is slightly less depressed. There is certainly very much scope to
    discuss the age of trimline, which may differ significantly from that of
    the maximum ice extent in the lowland, but we are not aware of any data
    that could be used for validation here. However, frozen areas plotted on
    Fig.~6c are now consistent with the modelled ages used in panels a and c.

@@

    \referee{%
        The text could briefly mention (and hopefully rule out) the issue of
        very fine-scale topographic features on which the trimlines are
        located, not resolved by the 1-km model topography. If data sites are
        on small-scale highs or lows significantly different from their
        ~km-scale surroundings, that could contribute to the discrepancies in
        Fig.~6c.}

    This is a valid point which we can unfortunately not rule out. Despite
    PISM's enthalpy scheme, which ensures a seamless transition between cold
    and temperate ice physics, resolving the basal thermal boundary on the
    valley sides is delicate, because it implies resolving a steep (vertical)
    enthalpy gradient over a steep bedrock slope. This requires both high
    vertical and horizontal grid resolutions. This limitation was mentioned
    in the section on trimlines:

    \msquote{%
        This result is certainly affected by the limited ability of the 1\,km
        horizontal resolution to resolve small-scale topographic features. It
        calls for a more detailed comparison.}

### 2018-07-20 articles/cycle/alps-cycle.tex
Adress referee 1 comments on trimlines.

    \begin{figure}[t]
      \centerline{\includegraphics{alpcyc_hr_trimlines}}
      \caption{%
        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
        (time-transgressive, corresponding to maximum ice thickness,
        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
        observed trimline elevations for the upper Rhone Glacier
        \citep[Table~1]{Kelly.etal.2004}. Colours
        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
        Model variables were bilinearly interpolated to the trimline locations.
        \textbf{(b)} Histogram of differences between modelled LGM ice surface
        elevation and trimline elevations (50\,m bands). The average
        difference is 861\,m.
        \textbf{(c)} Observed upper Rhone Valley trimline locations
        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
        (colour) and LGM ice surface elevation (200\,m contours). Hatches
        mark areas modelled to have been glaciated but to have experienced
        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).}
      \label{fig:trimlines}
    \end{figure}

@@

    In the upper Rhone Valley, the geographical location of observed trimlines
    show a remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, and to the uncertain subglacial
    topography the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

>>>>

    \begin{figure}[t]
      \centerline{\includegraphics{alpcyc_hr_trimlines}}
      \caption{%
        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
        (time-transgressive, corresponding to maximum ice thickness,
        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
        observed trimline elevations for the upper Rhone Glacier
        \citep[Table~1]{Kelly.etal.2004}. Colours
        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
        Model variables were bilinearly interpolated to the trimline locations.
        \textbf{(b)} Histogram of differences between modelled LGM ice surface
        elevation and trimline elevations (50\,m bands). The average
        difference is 861\,m.
        \textbf{(c)} Observed upper Rhone Valley trimline locations
        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
        (colour) and LGM ice surface elevation (200\,m contours). Hatches
        mark areas modelled to have been glaciated but to have experienced
        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).
        \todo{simplify basal thermal map, caption and main text.}}
      \label{fig:trimlines}
    \end{figure}

@@

    In the upper Rhone Valley, the geographical location of observed trimlines
    show a remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of
    the MIS~2 (Fig.~\ref{fig:trimlines}c). This result is certainly affected by
    the limited ability of the 1\,km horizontal resolution to resolve
    small-scale topographic features. It calls for a more detailed comparison
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.
